---
title: "Finding, importing and cleaning transport datasets"
---

# Introduction

In this session, we will explore how to find, download, import, and clean transport-related datasets. Transport data comes from various sources such as government agencies, open data portals, and crowd-sourced platforms. We'll learn practical techniques using R to access and prepare this data for analysis.

The session covers:

- Downloading datasets from OpenStreetMap
- Importing data into R
- Basic data cleaning and exploration
- Hands-on exercises

# Prerequisites

Before starting, ensure you have the necessary packages installed:

```{r}
#| output: false
options(repos = c(CRAN = "https://cloud.r-project.org"))
if (!require("pak")) install.packages("pak")
pkgs = c(
  "sf",
  "tidyverse",
  "osmextract",
  "tmap",
  "maptiles",
  "stats19",
  "pct"
)
pak::pkg_install(pkgs)
sapply(pkgs, require, character.only = TRUE)
```

## Project setup

Follow the guidance in the [day 1 slides](slides/day1.qmd#practical-session-0) to set up your project folder/repo.

# Downloading Transport Datasets

## OpenStreetMap Data

OpenStreetMap (OSM) provides global geographic data with a focus on human-made entities, including roads.
It is therefore very useful for quickly obtaining road network data for transport analysis.
A disadvantage of OSM data is that it can be inconsistent in quality and coverage, depending on the area, but for many applications these disadvantages are outweighed by the ease of access and free availability of the data.

Use the `osmextract` package to download and extract specific features.

```{r}
#| eval: false
library(osmextract)
# Download cycleways in West Yorkshire
west_yorkshire_cycleways = oe_get(
  # force_download = TRUE,
  place = "West Yorkshire",
  extra_tags = c("maxspeed", "lit", "cycleway"),
  query = "SELECT * FROM lines WHERE highway IN ('cycleway', 'path')"
)
plot(st_geometry(west_yorkshire_cycleways))
```

## Road Traffic Casualty Data (STATS19)

The UK's road traffic casualty data is available through the `stats19` package. This provides data on collisions, casualties, and vehicles.

```{r}
#| eval: false
library(stats19)
# Download 2020 collision data
collisions = get_stats19(year = 2020, type = "collision")
# Download casualty data
casualties = get_stats19(year = 2020, type = "cas")
# Download vehicle data
vehicles = get_stats19(year = 2020, type = "veh")
```

## Origin-Destination Data

The `pct` package provides access to the Propensity to Cycle Tool data, which includes origin-destination flows.

```{r}
#| eval: false
library(pct)
# Download desire lines for Leeds
leeds_desire_lines = get_pct_lines(region = "west-yorkshire")
```

## Boundary and Census Data

For geographic boundaries and census data, use the ONS Geoportal or custom dataset tool.

```{r}
#| eval: false
# Download Local Authority Districts
library(arcgis)
lad_url = "https://services1.arcgis.com/ESMARspQHYMw9BZ9/arcgis/rest/services/Local_Authority_Districts_May_2023_UK_BUC/FeatureServer/0"
lad_boundaries = arc_read(lad_url)
```

# Importing and Cleaning Data

Once downloaded, data often needs cleaning. Common issues include missing values, inconsistent formats, and coordinate systems.

## Basic Cleaning Example

```{r}
#| eval: false
# Clean collision data
collisions_clean = collisions |>
  # Remove rows with missing coordinates
  filter(!is.na(longitude), !is.na(latitude)) |>
  # Convert to sf object
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  # Select relevant columns
  select(accident_index, date, speed_limit, accident_severity)
```

## Handling Missing Data

```{r}
#| eval: false
# Check for missing values
summary(collisions_clean)
# Impute or remove missing values
collisions_clean = collisions_clean |>
  mutate(speed_limit = ifelse(is.na(speed_limit), median(speed_limit, na.rm = TRUE), speed_limit))
```

# Exercises

## Download and Explore STATS19 Data

1. Download road traffic collision data for 2019 using the `stats19` package.
2. Explore the structure of the data using `str()` and `summary()`.
3. Create a simple plot showing the number of collisions by severity.

```{r}
#| eval: false
# Your code here
```

## Extract OSM Features

1. Use `osmextract` to download all supermarkets in your chosen city.
2. Convert the data to an sf object and plot it on a map.
3. Count the number of supermarkets by type.

## Clean and Visualize OD Data

1. Download origin-destination data for a region using the `pct` package.
2. Clean the data by removing any invalid geometries.
3. Create a map showing the desire lines colored by cycling potential.

```{r}
#| eval: false
# Your code here
```

<!-- 
## Combine Datasets

1. Download boundary data for local authorities.
2. Download census data on travel to work patterns.
3. Join the datasets and create a choropleth map showing car commuting patterns.

```{r}
#| eval: false
# Your code here
```
-->

## Data Quality Assessment

1. Choose a dataset you've downloaded.
2. Assess its quality by checking for:
   - Missing values
   - Outliers
   - Inconsistent data types
3. Document any issues found and how you would address them.

```{r}
#| eval: false
# Your code here
```

# Further Reading

- [Reproducible Road Safety Research with R](https://itsleeds.github.io/rrsrr/)
- [Geocomputation with R](https://r.geocompx.org/)
- [Transport Data Science course materials](https://itsleeds.github.io/tds/)